Cartogram Mapping and its Application to Cancer Data Visualization

### 2019-12-08 version1/Response/Response.Rmd

---
from: Stephanie Kobakian
fromqualifications: BA
fromtitle: Research Masters Student
fromemail: stephanie.kobakian@monash.edu
fromphone: +61 433 699 797
toname: Professor Peter Baade and Dr Susanna Cramb
opening: Dear Professor Peter Baade and Dr Susanna Cramb
closing: Warm regards
spacing: 1.4
output: MonashEBSTemplates::letter
---

Thank you for your letter containing your feedback regarding our submision "Cartogram Mapping and its Application to Cancer Data Visualization", for inclusion in the Focused Issue on Spatial Patterns in Cancer Epidemiology in the Annals of Cancer Epidemiology.

This review paper presented an exploration of cancer mapping methods, focusing on variations on choropleths for online cancer atlases.

**Editor suggestion:**
**We would suggest that the aim of the manuscript could be phrased to highlight the current approaches and limitations of typical visualisation methods for cancer maps, along with outlining possible alternative methods and challenges in implementing these?**

We have restructed our original submission to provide a clear path from current approaches used, to the alternative approaches that should be explored. We highlight the limitations of using current display methods, to encourage the use of  alternative display methods. We also present the challenges that may be faced when implementing the alternative designs.

The manuscript now contains the following sections:

Current approaches
Limitations of current approaches
Alternative visualisation methods
Challenges of alternative approaches

We have inclued changes to reflect your minor notes:

**Editor suggestion:**
**The authors mention a focus on Atlases published between 2010 and 2015, but then include a bowel cancer atlas from 2016. **

**Also, within the bowel cancer atlas data, the reporting of percent of males aged 50-54 with bowel cancer in 2016 is incorrect; it should be the % of the ERP in each area that are males aged 50-54 years. **

We have corrected the description of the statistic used in the bowel cancer atlas of Australia:

Bowel Cancer Australia Atlas:
...shows the average Standardised Incidence Ratio of colorectal cancer for Australian males from 2006 to 2010 in Australia  . It is published by *[Bowel Cancer Australia](https://www.bowelcanceraustralia.org/)*.

**Reviewer A: **
**Major**
*Comment 1: *
The Abstract should be more focused- perhaps give an example of how cartogram mapping has been applied to data visualisation. What are the benefits of this approach and suggestions for the future.

*Reply 1:* 
We agree with the reviewer that alternative approaches have been fantastic when used in political contexts. Choropleth maps have often been used.

*Changes in the text:*

*Comment 2: *

2. The key message seems to get lost in the Abstract. The Conclusion is vague.

*Reply :*

*Changes in the text:*

*Comment 3: *

3. Entire paper needs substantial English language editing

*Reply :*

*Changes in the text:*

*Comment 4: *

4. Please avoid the use of sentences which have been used throughout the paper where you start with a reference number: such as “[4] discusses the use of (Map displays for disease data, Line 15-16) They do not make any sense and must be modified.

*Reply :*

*Changes in the text:*

*Comment 5: *

5. Please provide numbers for Figures and Tables

*Reply :*

*Changes in the text:*

*Comment 6: *

6. Please clearly state the primary aim of this paper and structure it accordingly.

*Reply :*

*Changes in the text:*

*Comment 7: *

7. Conclusions must discuss the key focus and conclusions of this paper.

*Reply :*

*Changes in the text:*

*Minor Comments*

- Please check the use of past and present tense
- There are many minor grammatical errors that need to checked.
- Please edit the last section of the Introduction. Either number each section and then refer to them or find another way to appropriately refer to each section.

We thank you for your consideration.

### 2019-12-08 review_article.Rmd
begin response

\section*{Abstract}
<!-- 119 words -->
Choropleth maps are widely used in cancer atlases, and traditionally communicate cancer statistics over geographic domains. The type of statistic displayed varies immensely, and effective choices can assist in the understanding of impact of cancer on a population. Atlases have moved from print media to web format, and the availability of graphical user interfaces, and user interaction and animation, greatly enhance the ability to communicate more information and positively engage readers. More use of alternatives to choropleth maps, such as cartograms, could be employed in online atlases, for better representation of spatial distributions, and show statistics on small areas. This is especially true for the geography of Australia, where the population is concentrated is small areas along the coastlines.

# Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Cancer statistics are usually delivered as an aggregated value for a geopolitical area. Presenting these statistics requires transforming individual observations into aggregations of communities as geographical units, in large part for privacy protection or political and policy purposes.  The information could be as simple as counts per area (e.g. state, province, local government area, post/zip code). Counts alone are not sufficient to compare areas, because the populations of areas are all different. In this case, the counts data needs to be merged with population data to appropriately calibrate it to incidence, for example, rate per 100000 people.  This type of data is collected on a routine basis for public health purposes, and may be made available to the general public as a service to the community. The task, then, is to examine what are the usual ways to communicate cancer statistics, to the public, are there alternative approaches, and what are the pros and cons of these choices.

<!-- Ways to visualize, whats our hypothesis -->
A common approach to communicate cancer statistics, is to display statistics on a map. Using a choropleth map: the statistic is mapped to color and the geographic region is filled with this color. The viewer would then be able to examine the spatial distribution of the disease incidence, where there is a trend in longitude or latitude, or rural vs urban, or coastal vs inland, or even specific hot spots of the disease. Visualizing diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses. Disease maps help to present geographic patterns that may be overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer outcomes, geographic patterns of disease may be identified and effectively addressed with public health policy and actions.
@SE recognizes one of the key challenges with mapping spatial patterns of disease is the design of visualizations.
This paper addresses the visualization techniques and their applications to cancer statistics. Highlighting the differences and historic use of these displays.

<!-- Structure of the paper -->
The paper is organized as follows. The next section describes the choropleth map which is the common approach to disease maps. Section \ref{ch:public} surveys atlases in use today. Section \ref{ch:alternatives} describes  alternative displays, including the cartogram which is useful when the map has heterogeneously sized geographic units. The pros and cons of these approaches is discussed in Section \ref{ch:compare}. Disease maps are more useful when made interactive, and common options are described in Section \ref{ch:interacting}, along with a discussion of benefits and disadvantages. The last section summarizes the paper.

# Map displays for disease data {#ch:choropleths}
<!-- 381 words -->
<!-- what is a choropleth map -->
A choropleth map is used to display differences in the geographical distribution of data by spatial unit by shading areas of a map [@EI]. The geography is faithfully rendered, and the color rendering is designed to reveal spatial patterns among data values. Identifying and explaining spatial structures, patterns, and processes involves considering the individuals in communities and organizing communities into representable units [@SAMGIS].
Early versions of choropleth maps used symbols or patterns instead of color.
@CPISACA discuss the use of choropleths to visualize cancer data, and @DMAHP gives an overview of the development of these maps for displaying disease data. In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality.

Figure \ref{fig:choroCRS} shows various geographic projections of choropleth maps of age-adjusted rate (per 100,000 people) of new cases of lung and bronchus in the USA, averaged over 2012 through 2016. The data was extracted from the official federal statistics by @USInteractive on cancer incidence and deaths, produced by the Centers for Disease Control and Prevention (CDC) and the National Cancer Institute (NCI). A west to east spatial trend of increasing rates, can be seen. There is also a spatial outlier -- Utah has a noticeably lower rate that its neighbors. Also Kentucky has a noticeably high rate, and Maine also has a higher rate than its neighbors. There is something of a cluster of higher rates around the tobacco states.

@@

# Cancer atlases {#ch:public}
<!-- 360 words -->
<!-- Displaying of cancer data on choropleths -->
A cancer atlas is a map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, geographic correlations, work related exposures, and high risk diets [@MACM]. The data collection of statistics across regions, and the administrative control within regions, lends itself to choropleth visualization. 
Atlases date to Haviland's maps in 1875, with UK cancer atlases appearing in 1961 [@HEDP], and early work in US cancer atlases appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power, and the availability of geographic information systems software [@SE].

<!-- Measures reported in cancer atlases -->
Mortality rates are now often presented as relative rates of risk across the population, and age adjusted to correct for the higher prevalence of cancers in older populations.
Howe [-@HEDP] describes Stock's development of the standardized mortality ratios through the 1930s. Table \ref{tab:measures} summarizes the measures commonly presented in published cancer atlases.

@@

|Measure |Details| 
|:--------------------|:------------------------------------------------|  
|1. Count|Crude cancer counts|
|2. Rate per 100,000|Cancer incidence per 100,000 population|  
|3. IR (Incidence Ratio)|$(IR)_i=\frac{(Incidence\ Rate)_i}{Average\ Incidence\ Rate}$, |
||Cancer incidence rate in region $i$ over the average cancer incidence rate for the total region|
|4. Age Adjusted Rate per 100,000| Standardized by age structure or region|   
|5. Age Adjusted Relative Risk| Standardized by age structure in each region $i$|  
|6. SIR (Standardized Incidence Ratio)| Standardized by incidence in each region $i$|
|7. Below or above Expected|Alternative expression of the SIR|
|8. RER |$RER = \frac{(Cancer\ related\ mortality)_i}{Average\ cancer\ related\ mortality}$|
|(Relative Excess Risk)|Represents the estimate of cancer related mortality within five years of diagnosis. Also referred to as 'excess hazard ratio'|

@@

Cancer maps are effective tools for communicating incidence, survival and mortality to wide range of audiences, including the general public and others not trained in statistical analyses.
These visualizations enable non-expert audiences to interpret the outputs of sophisticated statistical analyses. Cruickshank (1947) as cited by @DMAHP, discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, the cancer maps utilized to communicate to the public and other non-expert audiences are choropleths.

@@

## Overview of publicly available atlases \label{sec:publicatlases}
<!-- 416 words -->
Roberts [@roberts2019communication] identified 33 publicly available cancer atlases, published between January 2010 and November 2015, from different countries around the globe. All of these use choropleth maps. All but one of these were published by non-commercial organizations, including not-for-profits, government, research organizations, advocacy groups or government funded partnerships.  Figure \ref{fig:choropleth-grid-create} displays a subset of maps from these atlases.

@@

Figure \ref{fig:choropleth-grid-create}a shows the relative risk for women developing lung cancer in England and Wales in 2010 [@EnvEnglandWales2]. The cancer data was from Office for National Statistics (ONS) (England) and from the Welsh Cancer Intelligence and Surveillance Unit (WCISU).

@@

The map seen in Figure \ref{fig:choropleth-grid-create}b shows age standardized incidence rates (per 100,000) for all invasive cancers for both men and women, aggregated at a national level for 2018 [@Globocan]. It is published by the [World Health Organization's International Agency for Research on Cancer](https://www.iarc.fr/). Data was sourced from cancer registries of each country.

@@

### Bowel Cancer Australia Atlas 
Figure \ref{fig:choropleth-grid-create}d shows the percentage of Australian males between 50 - 54 years of age diagnosed with bowel cancer in 2016 in Australia [@Bowel]. It is published by *[Bowel Cancer Australia](https://www.bowelcanceraustralia.org/)*.

@@

Figure \ref{fig:choropleth-grid-create}f shows side by side maps of relative risk of lung cancer for men vs women for 2004 to 2008 [@cancerSpain].

@@

It is extremely likely that each population area will have a different number of people. The distribution of the population residing in all areas may also be communicated in a table or histogram display [@NICR_1].
Atlases can connect the population to the land available to them by communicating population density. Alternatives to choropleth to address this are described in Section \ref{ch:alternatives} .

@@

With any number comes some uncertainty about its value. This may occur for a number of reasons: numbers are drawn from samples, errors occur in the disease reporting, values are simulated from a model to maintain privacy. Close to half of the atlases identified (42%, n=14) included some measure of uncertainty. Uncertainty is often recorded as sample or population size, standard deviation, confidence intervals, statistical significance, quantiles, credible intervals. The most common measure used to represent uncertainty were credible or confidence intervals (CIs).

Displaying the uncertainty associated with reported statistics is a vital feature of a cancer map, but it can be difficult to display effectively. The map focuses on displaying the statistic, and lacks additional space to represent the uncertainty. Providing an adjacent map or overlaying maps with symbols [@VSSDCUC] are two approaches.

@@

Socio-economic indicators, such asunemployment rates, poverty rates, remoteness, and education levels,  can explain how the experience of cancer prevalence varies for various members of a society. Few atlases provide this level of detail. Data is available from the United Nations Human Development Index and financial indicators are available from the World Bank [@IARC_3].

@@

<!-- There are many alternative views of the world-->
<!-- Tobler map projections - I think this is not relevant here
Choropleths may be considered true topological maps, however, if the land mass displayed covers enough of the globe, there must be a transformation or distortion to display the land in 2D [@HTLWM]. The amount of distortion is related to the distance covered by the landmass displayed @GAMP.
World map projections reflect the frequent perspectives used to view the earth. Choropleth maps will always be distorted if they cover enough of the globe, just like photographs of the globe from space. Choropleth creation requires choosing a map projection that shows a favorable distortion of the geography for presenting the set of spatial information. Selecting a display can prevent misinterpretation of global statistics, as global maps face the challenge of equitable displays of land mass on maps [@RSCW].
If the statistic presented on the map base relies on physical distance and is influenced by the topology there is no transformation needed, beyond choosing a reasonable projection. 
-->

```{r choroCRS, fig.cap = "Four choropleth maps average age-adjusted rate of incidence for lung and bronchus in the United States averaged over sex and years 2012-2016, using four different map projections. The map projections alter the shapes and angles of the boundaries of each state.", message=FALSE, warning=FALSE, out.width='100%', fig.align='center'}
#knitr::include_graphics("figures/ggchoroCRS.png")
c1 <- png::readPNG("figures/gg3857.png")
c2 <- png::readPNG("figures/gg2163.png")
c3 <- png::readPNG("figures/gg4326.png")
c4 <- png::readPNG("figures/gg2955.png")
#knitr::include_graphics("figures/usa_grid.png")
load("figures/usa_legend.rda")

@@

<!-- Denominators -->
The purposeful distortion of the map space, transformed according to population density, is beneficial when a uniform density of the map base is desired. Population then becomes a uniformly distributed background for the statistic presented [@GOINO]. "population distribution is often extremely uneven in former British colonies" [@ACTUC ], and this makes the distortion necessary [@CTTMB].
When implementing a distortion of the geographical shape according to population, the resulting display is an area cartogram [@NAC], or population-by-area cartogram [@TAAM]. Event cartograms change the area of regions on a map depending on the amount of disease related events, but this does not consider the effects of land area and population [@VSSDCUC].

<!-- Why transform?-->
<!-- Common variables used to create cartograms, e.g. population, mortality -->
Cartograms provide an alternative visualization method for statistical and geographical information.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape or distance of geographical areas [@ACTUC]. 
Monmonier [@HTLWM] suggests that white lies may be employed to create useful displays and map creators have the ability to draw lines that may distort the geometry and suppress features and it is easy for the average person to disregard the impact of transformations used to create cartograms.
Cartograms may be seen as an extension of map transformations and projections. The favorable distortion is proportional to a value other other than actual earth size area [@NAC].
A disadvantage of the conventional map is that sparsely populated rural areas may be emphasized, whereas the areas representing cities are very small, making interpretation of spatial patterns very difficult.
The distortion of a cartogram accounts for the population density, preventing it from obscuring the spatial patterns [@TAAM]. 
The spatial transformation of map regions relative to the data emphasizes the data distribution instead of land size [@CBATCC].
When visualizing population statistics, Dorling [@ACTUC] considers this equitable representation design 'more socially just', or honest [@NISCC], giving due attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. 
Howe [@HEDP] suggests that 'cancer occurs in people, not in geographical areas' and @CTTMB believe that spatial socio-economic data, like cancer rates, are best presented on a cartogram for urban areas as the population map base avoids allocating 'undue prominence' to rural areas. 
<!-- @MTMSIH encourage the use of cartograms to highlight small areas and uncover local-level inequalities. DI SAYS I think that focuses on bayesian analysis not on cartograms-->

@@

The creation of cartograms was historically in the hands of professional cartographers [@CD]. 
Early approaches including John Hunter and Jonathan Young (1968) and Durham's wooden tile method, Skoda and Robertson's (1972) steel ball bearing approach and Tobler's (1973) computer programs [@ACTUC]. 
Geographical information systems allowed map users, and researchers to implement their own cartograms, but these systems are utilized depending on ‘the effectiveness, efficiency, and satisfaction of the map products (Nielsen 1994), [@CD].
Howe [@HEDP] discusses the impact of electronic computer-assisted techniques.

@@

```{r ggcartograms, fig.cap = "Common alternatives to maps, showing the same information as in Figure \ref{choroCRS} for the United States of America: (a) contiguous cartogram, (b) non-contiguous, shape-preserved cartogram, (c)  Dorling cartogram (non-contiguous), (d) hexmap (non-contiguous). In (a) - (c) the state has been resized, and reshaped, to match the 2015 population of the state. This provides a better sense of the extent of disease relative to the population in the country, and can help alleviate losing information about physically small but population dense states. In the hexmap (d) each state is given equal size, and thus equal emphasis.", message=FALSE, warning=FALSE, out.width='100%'}
g1 <- png::readPNG("figures/ggcont.png")
g2 <- png::readPNG("figures/ggncont.png")
g3 <- png::readPNG("figures/ggdorl.png")
g4 <- png::readPNG("figures/gghexmap.png")
#knitr::include_graphics("figures/usa_grid.png")

@@

|Map display |Details| 
|:--------------------|:------------------------------------------------|  
|a. Contiguous| Each stated shape has been distorted according to the population of the state in 2015. The state of California has become much larger due to it's large population density. This draws attention to the densely populated North East region, and detracts from the less populated Mid West.|
|b. Non-contiguous| The geographic shape of the states has been maintained, but the size has altered according to the population of the state in 2015. The state of California has remained closer to it's original size than it's surrounding states. The North East states have remained closer to their geographical size, in the case of Massachusetts and Connecticut. This draws attention to the densely populated North East region, and the sparse Mid West.|
|c. Dorling| The states have been represented by a circle, but the size was determined by the population of the state in 2015. The North East states remain closer to their neighbors, and may be displaced from their geographic location. The sparsity of the population in the Mid West is highlight by the distance between the circles, located at the geographic centroids.|
|d. Hexagon Tessellation| Each state is represented by a hexagon of equal size. The neighboring states are easily contrasted, however the north east regions has been displaced from their geographic location. The sparsity of the population in the Mid West is highlight by the light yellow color, the age adjusted rate in Kentucky is the darkest and its neighbors are similar.|

@@

Non-contiguous cartograms preserve shape of the area over connectivity. Each area stays in a similar position to their location on a choropleth map. The choropleth map base is often also plotted as a comparison point to highlight the change in area. The addition is the gap between areas, created as each individual area shrinks or grows according the associated value of the statistic. Olson [@NAC] discusses creation of these maps, the significance of the empty areas left between the geographic boundaries and the new shape, and the 'degree of difference from the original map that is the real message' of these displays.

@@

Daniel Dorling presented an alternative display engineers to highlight the spatial distribution and neighborhood relationships without complex distortions of borders and boundaries. This approach opposes preserving the intricate shape details and is founded in the simple question put forward by Daniel Dorling -@ACTUC:

>>>>

\section*{Abstract}
<!-- 119 words -->
<!--
Choropleth maps are widely used in cancer atlases to communicate cancer statistics over geographic domains. The type of statistic displayed varies immensely, and effective choices can assist in the understanding of impact of cancer on a population. Atlases have moved from print media to web format, and the availability of graphical user interfaces, and user interaction and animation, greatly enhance the ability to communicate more information and positively engage readers. More use of alternatives to choropleth maps, such as cartograms, could be employed in online atlases, for better representation of spatial distributions, and show statistics on small areas. This is especially true for the geography of Australia, where the population is concentrated is small areas along the coastlines.  
-->

Cancer atlases communicate cancer statistics over geographic domains. These domains are often subdivided by administrative areas such as countries, states or suburbs.
When communicating human-related statistics in Australia, the geographic map base may highlight the sparsely populated rural areas. The smaller geographic areas may not draw the attention of readers but they may be important to consider if they are more densely populated. To

# Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Cancer statistics are usually delivered as an aggregated value for a geopolitical area. Presenting these statistics requires transforming individual observations into aggregations of communities as geographical units, in large part for privacy protection or political and policy purposes.  The information could be as simple as counts per area (e.g. state, province, local government area, post/zip code). Counts alone are not sufficient to compare areas, because the populations of areas are all different. In this case, the count data needs to be merged with population data to appropriately calibrate it to incidence, for example, the rate per 100000 people.  This type of data is collected on a routine basis for public health purposes and may be made available to the general public as a service to the community. The task, then, is to examine what are the usual ways to communicate cancer statistics, to the public, are there alternative approaches, and what are the pros and cons of these choices.

<!-- Ways to visualize, whats our hypothesis -->
A common approach to communicate cancer statistics is to display statistics on a map. Using a choropleth map: the statistic is mapped to color and the geographic region is filled with this color. The viewer would then be able to examine the spatial distribution of the disease incidence, where there is a trend in longitude or latitude, or rural vs urban, or coastal vs inland, or even specific hot spots of the disease. Visualizing diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses. Disease maps help to present geographic patterns that may be overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer outcomes, geographic patterns of the disease may be identified and effectively addressed with public health policy and actions.
@SE recognizes one of the key challenges with mapping spatial patterns of disease is the design of visualizations.
This paper addresses the visualization techniques and their applications to cancer statistics. Highlighting the differences and historic use of these displays.

<!-- Structure of the paper -->
The paper is organized as follows. The next section describes the choropleth map which is the common approach to disease maps. Section \ref{ch:public} surveys atlases in use today. Section \ref{ch:alternatives} describes  alternative displays, including the cartogram which is useful when the map has heterogeneously sized geographic units. The pros and cons of these approaches are discussed in Section \ref{ch:compare}. Disease maps are more useful when made interactive, and common options are described in Section \ref{ch:interacting}, along with a discussion of benefits and disadvantages. The last section summarizes the paper.

# Map displays for disease data {#ch:choropleths}
<!-- 381 words -->
<!-- what is a choropleth map -->
A choropleth map is used to display differences in the geographical distribution of data by spatial unit by shading areas of a map [@EI]. The geography is faithfully rendered, and the color rendering is designed to reveal spatial patterns among data values. Identifying and explaining spatial structures, patterns, and processes involve considering the individuals in communities and organizing communities into representable units [@SAMGIS].
Early versions of choropleth maps used symbols or patterns instead of color.
@CPISACA discuss the use of choropleths to visualize cancer data, and @DMAHP gives an overview of the development of these maps for displaying disease data. In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality.

Figure \ref{fig:choroCRS} shows various geographic projections of choropleth maps of age-adjusted rate (per 100,000 people) of new cases of lung and bronchus in the USA, averaged from 2012 to 2016. The data was extracted from the official federal statistics by @USInteractive on cancer incidence and deaths, produced by the Centers for Disease Control and Prevention (CDC) and the National Cancer Institute (NCI). The rates exhibit a spatial trend, increasing from west to east. There is also a spatial outlier -- Utah has a noticeably lower rate than its neighbors. Also, Kentucky has a noticeably high rate, and Maine also has a higher rate than its neighbors. There is something of a cluster of higher rates around the tobacco states.

@@

# Cancer atlases {#ch:public}
<!-- 360 words -->
<!-- Displaying of cancer data on choropleths -->
A cancer atlas is a map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, geographic correlations, work-related exposures, and high-risk diets [@MACM]. The data collection of statistics across regions and the administrative control within regions lends itself to choropleth visualization. 
Atlases date to Haviland's maps in 1875, with UK cancer atlases appearing in 1961 [@HEDP], and early work in US cancer atlases appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power, and the availability of geographic information systems software [@SE].

<!-- Measures reported in cancer atlases -->
Mortality rates are now often presented as relative rates of risk across the population, and age-adjusted to correct for the higher prevalence of cancers in older populations.
Howe [-@HEDP] describes Stock's development of the standardized mortality ratios through the 1930s. Table \ref{tab:measures} summarizes the measures commonly presented in published cancer atlases.

@@

|Measure |Details| 
|:--------------------|:------------------------------------------------|  
|1. Count| Crude cancer counts|
|2. Rate per 100,000| Cancer incidence per 100,000 population|  
|3. IR (Incidence Ratio)| $(IR)_i=\frac{(Incidence\ Rate)_i}{Average\ Incidence\ Rate}$, |
|| The cancer incidence rate in region $i$ over the average cancer incidence rate for all of the regions|
|4. Age-Adjusted Rate per 100,000| Standardized by age structure or region|   
|5. Age-Adjusted Relative Risk| Standardized by age structure in each region $i$|  
|6. SIR (Standardized Incidence Ratio)| Standardized by incidence in each region $i$|
|7. Below or above Expected| An alternative expression of the SIR|
|8. RER | $RER = \frac{(Cancer\ related\ mortality)_i}{Average\ cancer\ related\ mortality}$|
|(Relative Excess Risk)| Represents the estimate of cancer-related mortality within five years of diagnosis. Also referred to as 'excess hazard ratio'|

@@

Cancer maps are effective tools for communicating incidence, survival, and mortality to a wide range of audiences, including the general public and others not trained in statistical analyses.
These visualizations enable non-expert audiences to interpret the outputs of sophisticated statistical analyses. Cruickshank (1947) as cited by @DMAHP, discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, the cancer maps utilized to communicate to the public and other non-expert audiences are choropleths.

@@

## Overview of publicly available atlases \label{sec:publicatlases}
<!-- 416 words -->
Roberts [@roberts2019communication] identified 33 publicly available cancer atlases, published between January 2010 and November 2015, from different countries around the globe. All of these use choropleth maps. All but one of these were published by non-commercial organizations, including not-for-profits, government, research organizations, advocacy groups or government-funded partnerships.  Figure \ref{fig:choropleth-grid-create} displays a subset of maps from these atlases.

@@

Figure \ref{fig:choropleth-grid-create}a shows the relative risk for women developing lung cancer in England and Wales in 2010 [@EnvEnglandWales2]. The cancer data was from the Office for National Statistics (ONS) (England) and the Welsh Cancer Intelligence and Surveillance Unit (WCISU).

@@

The map seen in Figure \ref{fig:choropleth-grid-create}b shows age-standardized incidence rates (per 100,000) for all invasive cancers for both men and women, aggregated at a national level for 2018 [@Globocan]. It is published by the [World Health Organization's International Agency for Research on Cancer](https://www.iarc.fr/). Data was sourced from cancer registries of each country.

@@

### Bowel Cancer Australia Atlas 
Figure \ref{fig:choropleth-grid-create}d shows the average Standardised Incidence Ratio of colorectal cancer for Australian males from 2006 to 2010 in Australia [@Bowel]. It is published by *[Bowel Cancer Australia](https://www.bowelcanceraustralia.org/)*.

@@

Figure \ref{fig:choropleth-grid-create}f shows side by side maps of the relative risk of lung cancer separately for men and women from 2004 to 2008 [@cancerSpain].

@@

Each population area will likely have a different number of people. The distribution of the population residing in all areas may also be communicated in a table or histogram display [@NICR_1].
Atlases can connect the population to the land available to them by communicating population density. Alternatives to choropleth to address this are described in Section \ref{ch:alternatives} .

@@

With any number comes some uncertainty about its value. This can occur for several reasons: numbers are drawn from samples, errors occur in the disease reporting, values are simulated from a model to maintain privacy. Close to half of the atlases identified (42%, n=14) included some measure of uncertainty. Uncertainty is often recorded as sample or population size, standard deviation, confidence intervals, statistical significance, quantiles, credible intervals. The most common measure used to represent uncertainty were credible or confidence intervals (CIs).

Displaying the uncertainty associated with reported statistics is a vital feature of a cancer map, but it can be difficult to display effectively. The map focuses on displaying the statistic and lacks additional space to represent the uncertainty. Providing an adjacent map or overlaying maps with symbols [@VSSDCUC] are two approaches.

@@

Socio-economic indicators, such as unemployment rates, poverty rates, remoteness, and education levels,  can explain how the experience of cancer prevalence varies for various members of society. Few atlases provide this level of detail. Data is available from the United Nations Human Development Index and financial indicators are available from the World Bank [@IARC_3].

@@

<!-- There are many alternative views of the world-->
<!-- Tobler map projections - I think this is not relevant here
Choropleths may be considered true topological maps however,, if the landmass displayed covers enough of the globe, there must be a transformation or distortion to display the land in 2D [@HTLWM]. The amount of distortion is related to the distance covered by the landmass displayed @GAMP.
World map projections reflect the frequent perspectives used to view the earth. Choropleth maps will always be distorted if they cover enough of the globe, just like photographs of the globe from space. Choropleth creation requires choosing a map projection that shows a favorable distortion of the geography for presenting the set of spatial information. Selecting a display can prevent misinterpretation of global statistics, as global maps face the challenge of equitable displays of landmass on maps [@RSCW].
If the statistic presented on the map base relies on physical distance and is influenced by the topology there is no transformation needed, beyond choosing a reasonable projection. 
-->

```{r choroCRS, fig.cap = "Four choropleth maps average age-adjusted rate of incidence for lung and bronchus in the United States averaged from sex and years 2012-2016, using four different map projections. The map projections alter the shapes and angles of the boundaries of each state.", message=FALSE, warning=FALSE, out.width='100%', fig.align='center'}
#knitr::include_graphics("figures/ggchoroCRS.png")
c1 <- png::readPNG("figures/gg3857.png")
c2 <- png::readPNG("figures/gg2163.png")
c3 <- png::readPNG("figures/gg4326.png")
c4 <- png::readPNG("figures/gg2955.png")
#knitr::include_graphics("figures/usa_grid.png")
load("figures/usa_legend.rda")

@@

<!-- Denominators -->
The purposeful distortion of the map space, transformed according to population density, is beneficial when a uniform density of the map base is desired. The population then becomes a uniformly distributed background for the statistic presented [@GOINO]. "population distribution is often extremely uneven in former British colonies" [@ACTUC ], and this makes the distortion necessary [@CTTMB].
When implementing a distortion of the geographical shape according to population, the resulting display is an area cartogram [@NAC], or population-by-area cartogram [@TAAM]. Event cartograms change the area of regions on a map depending on the amount of disease-related events, but this does not consider the effects of land area and population [@VSSDCUC].

<!-- Why transform?-->
<!-- Common variables used to create cartograms, e.g. population, mortality -->
Cartograms provide an alternative visualization method for statistical and geographical information.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape or distance of geographical areas [@ACTUC]. 
Monmonier [@HTLWM] suggests that white lies may be employed to create useful displays and map creators can draw lines that may distort the geometry and suppress features and it is easy for the average person to disregard the impact of transformations used to create cartograms.
Cartograms may be seen as an extension of map transformations and projections. The favorable distortion is proportional to a valueother than the actual earth size area [@NAC].
A disadvantage of the conventional map is that sparsely populated rural areas may be emphasized, whereas the areas representing cities are very small, interpreting spatial patterns very difficult.
The distortion of a cartogram accounts for the population density, preventing it from obscuring the spatial patterns [@TAAM]. 
The spatial transformation of map regions relative to the data emphasizes the data distribution instead of land size [@CBATCC].
When visualizing population statistics, Dorling [@ACTUC] considers this equitable representation design 'more socially just', or honest [@NISCC], giving due attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. 
Howe [@HEDP] suggests that 'cancer occurs in people, not in geographical areas' and @CTTMB believe that spatial socio-economic data, like cancer rates, are best presented on a cartogram for urban areas as the population map base avoids allocating 'undue prominence' to rural areas. 
<!-- @MTMSIH encourage the use of cartograms to highlight small areas and uncover local-level inequalities. DI SAYS I think that focuses on bayesian analysis not on cartograms-->

@@

The creation of cartograms was historically in the hands of professional cartographers [@CD]. 
Early approaches including John Hunter and Jonathan Young (1968) and Durham's wooden tile method, Skoda and Robertson's (1972) steel ball-bearing approach and Tobler's (1973) computer programs [@ACTUC]. 
Geographical information systems allowed map users, and researchers to create cartograms, but these systems are utilized depending on ‘the effectiveness, efficiency, and satisfaction of the map products (Nielsen 1994), [@CD].
Howe [@HEDP] discusses the impact of electronic computer-assisted techniques.

@@

```{r ggcartograms, fig.cap = "Common alternatives to maps, showing the same information as in Figure \ref{choroCRS} for the United States of America: (a) contiguous cartogram, (b) non-contiguous, shape-preserved cartogram, (c)  Dorling cartogram (non-contiguous), (d) hexagon tilemap (non-contiguous). In (a) - (c) the state has been resized, and reshaped, to match the 2015 population of the state. This provides a better sense of the extent of disease relative to the population in the country and can help alleviate losing information about physically small but population-dense states. In the hexagon tilemap (d) each state is given equal size and thus equal emphasis.", message=FALSE, warning=FALSE, out.width='100%'}
g1 <- png::readPNG("figures/ggcont.png")
g2 <- png::readPNG("figures/ggncont.png")
g3 <- png::readPNG("figures/ggdorl.png")
g4 <- png::readPNG("figures/gghexmap.png")
#knitr::include_graphics("figures/usa_grid.png")

@@

|Map display |Details| 
|:--------------------|:------------------------------------------------|  
|a. Contiguous| Each stated shape has been distorted according to the population of the state in 2015. The state of California has become much larger due to its large population density. This draws attention to the densely populated North-East region and detracts from the less populated Mid West.|
|b. Non-contiguous| The geographic shape of the states has been maintained, but the size has altered according to the population of the state in 2015. The state of California has remained closer to its original size than it's surrounding states. The North-East states have remained closer to their geographical size, in the case of Massachusetts and Connecticut. This draws attention to the densely populated North-East region and the sparse Mid West.|
|c. Dorling| The states have been represented by a circle, but the size was determined by the population of the state in 2015. The North-East states remain closer to their neighbors and could be displaced from their geographic location. The sparsity of the population in the Mid West is highlighted by the distance between the circles, located at the geographic centroids.|
|d. Hexagon Tessellation| Each state is represented by a hexagon of equal size. The neighboring states are easily contrasted however the North-East regions have been displaced from their geographic location. The sparsity of the population in the Mid West is highlighteded by the light yellow color, the Age-Adjusted rate in Kentucky is the darkest and its neighbors are similar.|

@@

Non-contiguous cartograms prioritize the shapes of the areas instead of connectivity. Each area stays in a similar position to its location on a choropleth map. The choropleth map base is often also displayed to allow comparisons to be made regarding the change in the area displayed. The addition is the gap between areas, created as each area shrinks or grows according to the associated value of the statistic. Olson [@NAC] discusses the creation of these maps, the significance of the empty areas left between the geographic boundaries and the new shape, and the 'degree of difference from the original map that is the real message' of these displays.

@@

Daniel Dorling presents an alternative display engineered to highlight the spatial distribution and neighborhood relationships without complex distortions of borders and boundaries. This approach opposes preserving the intricate shape details and is founded in the simple question put forward by Daniel Dorling -@ACTUC:
